---
title: "Problem Set 5 - Solutions"
format: pdf
---

# Problem 1

Consider a monopoly seller with zero production cost who chooses a price to maximise profits (revenue). The consumers have one of two valuations $v \in \{1, 2\}$ with $\mathbb{P}(v = 2) = x$.

a. Suppose that the monopolist has no additional information about the valuation beyond her prior belief $x$. What is the optimal price (as a function of $x$)? What is the expected profit and consumer surplus?
b. Suppose that the monopolist receives perfect information about the valuation. What are the optimal prices? What is the expected profit and consumer surplus?
c. Plot the (consumer surplus, profit) pairs from parts (a) and (b) on the same diagram assuming $x = 0.4$.
d. Plot the (consumer surplus, profit) pairs from parts (a) and (b) on the same diagram assuming $x = 0.6$.
e. Plot the consumer surplus from part (a) as a function of $x$.
f. Suppose that the consumers can commit to revealing any information about their valuation to the monopolist. What is the maximum consumer surplus that can be achieved when $x = 0.4$? What about when $x = 0.6$?
g. Assuming $x = 0.6$, add to your diagram from part (d) the (consumer surplus, profit) pair that maximises consumer surplus. Hence, find all (consumer surplus, profit) pairs that can result from some information structure for the monopolist.

# Solution to Problem 1

## Part (a): No additional information

The monopolist sets a single price $p$. Since valuations are $v \in \{1, 2\}$, the only
candidate prices are $p = 1$ and $p = 2$ (any other price is weakly dominated by
one of these).

- If $p = 1$: Both profiles buy. Profit $\pi= 1$. Consumer surplus: profile 1 gets
  $1 - 1 = 0$, profile 2 gets $2 - 1 = 1$. Expected CS $= x \cdot 1 = x$.
- If $p = 2$: Only profile 2 buys. Profit $= 2x$. Consumer surplus: profile 2
  gets $2 - 2 = 0$, profile 1 doesn't buy. Expected CS $= 0$.

The monopolist chooses the price that gives higher profit:
 
$$
p^*(x) =
\begin{cases}
1 & \text{if } x \leq \frac{1}{2} \\
2 & \text{if } x > \frac{1}{2}
\end{cases}
$$

$$
\pi(x) = \max(1,\; 2x) =
\begin{cases}
1 & \text{if } x \leq \frac{1}{2} \\
2x & \text{if } x > \frac{1}{2}
\end{cases}
$$

$$
\text{CS}(x) =
\begin{cases}
x & \text{if } x \leq \frac{1}{2} \\
0 & \text{if } x > \frac{1}{2}
\end{cases}
$$

## Part (b): Perfect information

If the monopolist knows $v$ exactly, she charges $p = v$:

- If $v = 1$: charges $p = 1$, profit $= 1$, CS $= 0$.
- If $v = 2$: charges $p = 2$, profit $= 2$, CS $= 0$.

Expected profit $= (1 - x) \cdot 1 + x \cdot 2 = 1 + x$.

Expected CS $= 0$ (the monopolist captures the entire surplus).

## Part (c): Diagram for $x = 0.4$

From part (a) with $x = 0.4$: since $0.4 < 1/2$, the optimal price is $p = 1$.
Profit $= 1$, CS $= 0.4$. Point: $(\text{CS}, \pi) = (0.4,\; 1)$.

From part (b) with $x = 0.4$: profit $= 1.4$, CS $= 0$.
Point: $(\text{CS}, \pi) = (0,\; 1.4)$.

```{r}
#| label: part-c
#| fig-cap: "(CS, Profit) pairs for x = 0.4"
#| warning: false
#| message: false

library(ggplot2)

dpoints_c <- data.frame(
  cs = c(0.4, 0),
  profit = c(1, 1.4),
  label = c("No information\n(0.4, 1)", "Perfect information\n(0, 1.4)")
)

ggplot(dpoints_c, aes(x = cs, y = profit)) +
  geom_point(size = 3) +
  geom_text(aes(label = label), vjust = -0.8, size = 3.5) +
  scale_x_continuous(
    "Consumer Surplus",
    limits = c(-0.1, 0.6),
    breaks = seq(0, 0.6, 0.1)
  ) +
  scale_y_continuous(
    "Profit",
    limits = c(0.8, 1.6),
    breaks = seq(0.8, 1.6, 0.2)
  ) +
  theme_minimal() +
  ggtitle("x = 0.4")
```

## Part (d): Diagram for $x = 0.6$

From part (a) with $x = 0.6$: since $0.6 > 1/2$, the optimal price is $p = 2$.
Profit $= 1.2$, CS $= 0$. Point: $(\text{CS}, \pi) = (0,\; 1.2)$.

From part (b) with $x = 0.6$: profit $= 1.6$, CS $= 0$.
Point: $(\text{CS}, \pi) = (0,\; 1.6)$.

Note that both points lie on the vertical axis — when $x > 1/2$ the monopolist
prices high even without information, so consumers get zero surplus either way.

```{r}
#| label: part-d
#| fig-cap: "(CS, Profit) pairs for x = 0.6"
#| warning: false
#| message: false

dpoints_d <- data.frame(
  cs = c(0, 0),
  profit = c(1.2, 1.6),
  label = c("No information\n(0, 1.2)", "Perfect information\n(0, 1.6)")
)

ggplot(dpoints_d, aes(x = cs, y = profit)) +
  geom_point(size = 3) +
  geom_text(aes(label = label), hjust = -0.15, size = 3.5) +
  scale_x_continuous(
    "Consumer Surplus",
    limits = c(-0.05, 0.6),
    breaks = seq(0, 0.6, 0.1)
  ) +
  scale_y_continuous(
    "Profit",
    limits = c(1, 1.8),
    breaks = seq(1, 1.8, 0.2)
  ) +
  theme_minimal() +
  ggtitle("x = 0.6")
```

## Part (e): Consumer surplus from part (a) as a function of $x$

From part (a), CS$(x) = x$ for $x \leq 1/2$ and CS$(x) = 0$ for $x > 1/2$.

```{r}
#| label: part-e
#| fig-cap: "Consumer surplus as a function of x (no information)"
#| warning: false
#| message: false

dcs <- data.frame(
  x = c(0, 0.5, 0.5, 1),
  cs = c(0, 0.5, 0, 0)
)

ggplot(dcs, aes(x = x, y = cs)) +
  geom_line(linewidth = 1) +
  geom_point(
    data = data.frame(x = 0.5, cs = 0.5),
    size = 3, shape = 16
  ) +
  geom_point(
    data = data.frame(x = 0.5, cs = 0),
    size = 3, shape = 1
  ) +
  scale_x_continuous("x = P(v = 2)", breaks = seq(0, 1, 0.1)) +
  scale_y_continuous("Consumer Surplus", breaks = seq(0, 0.5, 0.1)) +
  theme_minimal() +
  ggtitle("CS(x) under no information")
```

The consumer surplus rises linearly up to $x = 1/2$ (where at the optimum
$p = 1$, the high types enjoy surplus $2 - 1 = 1$) and
then drops to zero at $x = 1/2$ once the monopolist switches to $p = 2$.

## Part (f): Maximum consumer surplus with information design

The consumers choose an information structure (signal about $v$) that the
monopolist observes before pricing. The monopolist best-responds to her
posterior belief $\mu$. The constraint is Bayes plausibility: the expected
posterior must equal the prior, $\mathbb{E}[\mu] = x$.

Consumer surplus as a function of the posterior $\mu$ is:

$$
\text{CS}(\mu) =
\begin{cases}
\mu & \text{if } \mu \leq \frac{1}{2} \quad (p = 1,\; \text{high types get surplus } 1) \\
0 & \text{if } \mu > \frac{1}{2} \quad (p = 2,\; \text{no surplus})
\end{cases}
$$

To find the maximum expected CS over all information structures, we compute the
concavification (smallest concave function $\geq$ CS) of CS$(\mu)$:

$$
\text{CS}^{\text{cav}}(\mu) =
\begin{cases}
\mu & \text{if } \mu \leq \frac{1}{2} \\
1 - \mu & \text{if } \mu > \frac{1}{2}
\end{cases}
$$

This is the line connecting $(1/2, 1/2)$ to $(1, 0)$ for $\mu > 1/2$, which lies
above CS $= 0$ on that region.

When $x = 0.4$: $\text{CS}^{\text{cav}}(0.4) = 0.4$. This equals CS$(0.4)$
from part (a), so no information design can improve on revealing nothing.
Maximum CS $= 0.4$.

This makes sense: when $x < 1/2$, the monopolist already prices at $p = 1$ and
serves everyone. Any information that shifts the posterior above $1/2$ would
cause the monopolist to raise the price, hurting consumers.

When $x = 0.6$: $\text{CS}^{\text{cav}}(0.6) = 1 - 0.6 = 0.4$. Maximum CS
$= 0.4$ (a large improvement over CS $= 0$ with no information).

### Optimal information structure for $x = 0.6$

Split the prior into posteriors $\mu_L = 1/2$ and $\mu_H = 1$:

$$
\lambda \cdot \frac{1}{2} + (1 - \lambda) \cdot 1 = 0.6 \implies \lambda = 0.8
$$

- With probability 0.8: posterior is $1/2$, monopolist charges $p = 1$, CS $= 1/2$.
- With probability 0.2: posterior is $1$, monopolist charges $p = 2$, CS $= 0$.

Expected CS $= 0.8 \times 0.5 = 0.4$. Expected profit $= 0.8 \times 1 + 0.2 \times 2 = 1.2$.

Concrete signal structure: If the consumer has $v = 1$, always send signal
$s_L$. If $v = 2$, send $s_L$ with probability $2/3$ and $s_H$ with probability
$1/3$. The monopolist's posterior after $s_L$ is $1/2$ (charges $p = 1$) and
after $s_H$ is $1$ (charges $p = 2$).

## Part (g): Full set of achievable (CS, profit) pairs for $x = 0.6$

From part (f), the max-CS point is $(\text{CS}, \pi) = (0.4,\; 1.2)$.

We now characterise all (CS, profit) pairs achievable by some information
structure.

Key observations:

1. Minimum profit is $\pi(x) = 1.2$: The profit function
   $\pi(\mu) = \max(1, 2\mu)$ is convex, so by Jensen's inequality any
   information structure satisfies
   $\mathbb{E}[\pi(\mu)] \geq \pi(\mathbb{E}[\mu]) = 1.2$.

2. Maximum profit is $1 + x = 1.6$: achieved by perfect information.

3. When $\mu_L = 1/2$ and $\mu_H \in (1/2, 1]$: profit is always $1.2$ and
   CS ranges from $0$ to $0.4$. This gives the bottom edge of the feasible
   set.

4. When $\mu_L \in [0, 1/2]$ and $\mu_H = 1$: total realised surplus is
   constant at $1.6$ (everyone trades). So CS $+ \pi = 1.6$, giving the
   upper edge with slope $-1$.

The feasible set is a triangle with vertices:

| Point | CS | Profit | Description |
|---|---|---|---|
| A | 0 | 1.2 | No information |
| B | 0 | 1.6 | Perfect information |
| C | 0.4 | 1.2 | Max-CS information |

```{r}
#| label: part-g
#| fig-cap: "All achievable (CS, Profit) pairs for x = 0.6"
#| warning: false
#| message: false

# Triangle vertices
dtriangle <- data.frame(
  cs = c(0, 0, 0.4, 0),
  profit = c(1.2, 1.6, 1.2, 1.2)
)

# Labelled points
dpoints_g <- data.frame(
  cs = c(0, 0, 0.4),
  profit = c(1.2, 1.6, 1.2),
  label = c(
    "A: No info\n(0, 1.2)",
    "B: Perfect info\n(0, 1.6)",
    "C: Max CS\n(0.4, 1.2)"
  )
)

ggplot() +
  geom_polygon(
    data = dtriangle, aes(x = cs, y = profit),
    fill = "steelblue", alpha = 0.25
  ) +
  geom_path(
    data = dtriangle, aes(x = cs, y = profit),
    linewidth = 0.8
  ) +
  geom_point(data = dpoints_g, aes(x = cs, y = profit), size = 3) +
  geom_text(
    data = dpoints_g, aes(x = cs, y = profit, label = label),
    vjust = c(1.5, -0.5, -0.5), hjust = c(-0.1, -0.1, 0.5),
    size = 3.5
  ) +
  annotate("text", x = 0.08, y = 1.38,
    label = "Feasible set", fontface = "italic", size = 4) +
  scale_x_continuous(
    "Consumer Surplus",
    limits = c(-0.05, 0.55),
    breaks = seq(0, 0.5, 0.1)
  ) +
  scale_y_continuous(
    "Profit",
    limits = c(1.0, 1.8),
    breaks = seq(1.0, 1.8, 0.2)
  ) +
  theme_minimal() +
  ggtitle("x = 0.6: All (CS, Profit) pairs from information structures")
```

The shaded triangle is the set of all (consumer surplus, profit) pairs that can
result from some information structure. Any point inside or on the boundary is
achievable.

- Edge A–B (left): CS $= 0$, profit $\in [1.2, 1.6]$. Achieved by
  information structures where the monopolist always prices at 2 (both posteriors
  above $1/2$, or one at 0 and one above $1/2$).
- Edge B–C (top): CS $+ \pi = 1.6$ (no deadweight loss). Achieved by
  splitting the prior into $\mu_L \in [0, 1/2]$ and $\mu_H = 1$.
- Edge A–C (bottom): $\pi = 1.2$ (minimum profit). Achieved by splitting
  into $\mu_L = 1/2$ and $\mu_H \in (1/2, 1]$.

