---
title: "Problem Set 6 - Solutions"
format: pdf
---

# Problem 1

Consider a monopoly seller with zero production cost who chooses a price to maximise
profits (revenue). The buyer has one of two valuations $v \in \{1, 2\}$ with $\Pr(v = 2) = 0.6$.
The seller has no information beyond her prior belief. We will consider different information structures for the buyer.

a. Suppose the buyer has no information about her valuation. What price is optimal for
   the monopolist to set? What is the maximised profit? What is the consumer surplus?
b. Suppose that the buyer has perfect information about her valuation. What price is
   optimal for the monopolist to set? What is the maximised profit? What is the consumer surplus?

Let $F$ be the cdf of $v$, i.e.

$$
F(x) = \begin{cases}
0 & \text{if } x < 1 \\
0.4 & \text{if } x \in [1, 2) \\
1 & \text{if } x \geq 2
\end{cases}
$$

Suppose that the buyer receives a signal $s$ about her valuation, where $s = E(v|s)$.

c. Which cdfs $G$ of $s$ are consistent with $s = E(v|s)$?
d. Suppose the seller sets price $x$. What is the buyer's decision as a function of $s$?
e. Suppose that the seller is indifferent between setting price 1 and $x$. What must $G(x)$ be?
f. Suppose that

$$
G(x) = \begin{cases}
0 & \text{if } x < 1 \\
1 - \frac{1}{x} & \text{if } x \in [1, \bar{x}) \\
1 & \text{if } x \geq \bar{x}
\end{cases}
$$

where $\bar{x} \in (1, 2)$. Plot the graphs of $F$ and $G$ on the same diagram.

g. Find $\bar{x}$ such that $F$ is a mean preserving spread of $G$.
h. Given the $\bar{x}$ found in (g), what is the consumer surplus and profit?
i. Plot all the possible (consumer surplus, profit) pairs that can arise from some information structure for the buyer.

# Solution to Problem 1

This problem explores buyer information design—how different information structures
available to the buyer affect market outcomes. The key insight is that less informed
buyers are easier to exploit, while perfectly informed buyers create standard adverse
selection problems.

*Setup:* Seller has zero cost, buyer has $v \in \{1, 2\}$ with $\Pr(v = 2) = 0.6$,
$\Pr(v = 1) = 0.4$. Expected value is $E[v] = 0.4 \times 1 + 0.6 \times 2 = 1.6$.

## Part (a): Buyer has no information

When the buyer doesn't know her own valuation, she only knows $E[v] = 1.6$.

*Buyer's decision:* Buy if and only if $p \leq E[v] = 1.6$.

*Seller's optimization:* The seller can set any price $p \leq 1.6$ and the buyer will purchase. Revenue is maximized at $p^* = 1.6$.

$$
\text{Profit} = 1.6
$$

$$
\text{Consumer Surplus} = E[v - p] = E[v] - p = 1.6 - 1.6 = 0
$$

*Intuition:* An uninformed buyer reveals nothing through her purchasing decision, so the seller can extract the full expected surplus.

## Part (b): Buyer has perfect information

When the buyer knows her type exactly:

- Type $v = 1$ buys if and only if $p \leq 1$
- Type $v = 2$ buys if and only if $p \leq 2$

*Seller's options:*

- $p = 1$: Both types buy. Profit $= 1$.
- $p = 2$: Only $v = 2$ buys. Profit $= 0.6 \times 2 = 1.2$.

Since $1.2 > 1$, the optimal price is $p^* = 2$.

$$
\text{Profit} = 1.2
$$

$$
\text{Consumer Surplus} = 0.4 \times 0 + 0.6 \times (2 - 2) = 0
$$

*Intuition:* With perfect buyer information, the seller faces adverse selection. She sets a high price, excluding low types, but still extracts all surplus from high types.

```{r}
#| label: part-ab
#| fig-cap: "Comparison of no information vs perfect information"
#| warning: false
#| message: false

library(ggplot2)

dpoints_ab <- data.frame(
  cs = c(0, 0),
  profit = c(1.6, 1.2),
  label = c("No buyer info\n(0, 1.6)", "Perfect buyer info\n(0, 1.2)")
)

ggplot(dpoints_ab, aes(x = cs, y = profit)) +
  geom_point(size = 3) +
  geom_text(aes(label = label), hjust = -0.15, size = 3.5) +
  scale_x_continuous(
    "Consumer Surplus",
    limits = c(-0.05, 0.8),
    breaks = seq(0, 0.8, 0.2)
  ) +
  scale_y_continuous(
    "Profit",
    limits = c(0.9, 1.8),
    breaks = seq(1, 1.8, 0.2)
  ) +
  theme_minimal() +
  ggtitle("Parts (a) and (b): Extreme information structures")
```

## Part (c): Consistent CDFs for signal $s = E(v|s)$

A CDF $G$ is consistent with $s = E(v|s)$ if and only if:

1. *Support in $[1, 2]$:* Since $v \in \{1, 2\}$, any conditional expectation $E[v|s]$ must lie in $[1, 2]$.

2. *Bayes plausibility:* By the law of iterated expectations,
$$
E[s] = E[E(v|s)] = E[v] = 1.6
$$

So $G$ must have mean 1.6.

3. *Mean-preserving contraction:* $G$ must be a mean-preserving contraction of $F$. Equivalently, $F$ must be a mean-preserving spread of $G$.

Any distribution $G$ on $[1, 2]$ with mean 1.6 that is a mean-preserving contraction of $F$ is consistent.

## Part (d): Buyer's decision given signal $s$

The buyer observes signal $s = E(v|s)$, which is her expected valuation conditional on the signal. Given price $x$:

$$
\text{Buyer purchases} \iff s \geq x
$$

The buyer buys whenever her expected value (given her information) exceeds the price.

## Part (e): Seller indifference condition

The seller's profit at price $p$ is:
$$
\pi(p) = p \cdot \Pr(s \geq p) = p \cdot (1 - G(p))
$$

At price $p = 1$: $\pi(1) = 1 \cdot (1 - G(1)) = 1 \cdot 1 = 1$ (since $G(1) = 0$).

For indifference between price 1 and price $x$:
$$
1 = x \cdot (1 - G(x))
$$

Solving for $G(x)$:
$$
G(x) = 1 - \frac{1}{x}
$$

## Part (f): Plot of $F$ and $G$

```{r}
#| label: part-f
#| fig-cap: "CDFs F (valuation) and G (signal) for x̄ = e^0.6"
#| warning: false
#| message: false

xbar <- exp(0.6)

# Create data for F (step function)
x_vals <- seq(0, 2.5, length.out = 500)
F_vals <- ifelse(x_vals < 1, 0, ifelse(x_vals < 2, 0.4, 1))

# Create data for G
G_vals <- ifelse(x_vals < 1, 0, ifelse(x_vals < xbar, 1 - 1/x_vals, 1))

df_F <- data.frame(x = x_vals, cdf = F_vals, Distribution = "F (valuation)")
df_G <- data.frame(x = x_vals, cdf = G_vals, Distribution = "G (signal)")
df_both <- rbind(df_F, df_G)

ggplot(df_both, aes(x = x, y = cdf, color = Distribution,
    linetype = Distribution)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = xbar, linetype = "dashed", alpha = 0.5) +
  annotate("text", x = xbar + 0.05, y = 0.5, 
           label = paste0("x̄ = e^0.6 ≈ ", round(xbar, 3)), 
           hjust = 0, size = 3.5) +
  scale_x_continuous("x", limits = c(0, 2.5), breaks = c(0, 1, xbar, 2), 
                     labels = c("0", "1", "x̄", "2")) +
  scale_y_continuous("CDF", breaks = seq(0, 1, 0.2)) +
  scale_color_manual(values = c("F (valuation)" = "blue",
    "G (signal)" = "red")) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  ggtitle("F and G on the same diagram")
```

## Part (g): Finding $\bar{x}$ for mean-preserving spread

For $F$ to be a mean-preserving spread of $G$, we need $E_G[s] = E_F[v] = 1.6$.

The distribution $G$ has:
- Density $g(x) = \frac{d}{dx}(1 - 1/x) = \frac{1}{x^2}$ for $x \in [1, \bar{x})$
- Point mass of size $1 - G(\bar{x}^-) = 1 - (1 - 1/\bar{x}) = \frac{1}{\bar{x}}$ at $x = \bar{x}$

Computing $E[s]$:
$$
E[s] = \int_1^{\bar{x}} x \cdot \frac{1}{x^2} dx + \bar{x} \cdot \frac{1}{\bar{x}}
$$
$$
= \int_1^{\bar{x}} \frac{1}{x} dx + 1 = \ln(\bar{x}) + 1
$$

Setting $E[s] = 1.6$:
$$
\ln(\bar{x}) + 1 = 1.6 \implies \ln(\bar{x}) = 0.6 \implies \bar{x} = e^{0.6} \approx 1.822
$$

## Part (h): Consumer surplus and profit at $\bar{x} = e^{0.6}$

With this information structure, the seller is indifferent between all prices $p \in [1, \bar{x}]$, each yielding:

$$
\text{Profit} = 1
$$

The consumer surplus depends on which price the seller sets:

*If $p = 1$:* All buyers purchase.
$$
\text{CS} = E[s - 1] = E[s] - 1 = 1.6 - 1 = 0.6
$$

*If $p = \bar{x}$:* Only buyers with $s = \bar{x}$ (the point mass) purchase.
$$
\text{CS} = (\bar{x} - \bar{x}) \cdot \frac{1}{\bar{x}} = 0
$$

So with profit $= 1$, consumer surplus can range from $0$ to $0.6$.

```{r}
#| label: part-h
#| fig-cap: "Profit indifference at different prices"
#| warning: false
#| message: false

xbar <- exp(0.6)
prices <- seq(1, xbar, length.out = 100)

# At each price, profit = 1 (by construction)
# CS = E[s - p | s >= p] * Pr(s >= p)
# For s with density 1/x^2 on [1, xbar) and atom 1/xbar at xbar:

compute_cs <- function(p) {
  if (p >= xbar) return(0)
  # Integral from p to xbar of (s - p) * (1/s^2) ds + (xbar - p) * (1/xbar)
  integral_part <- integrate(
    function(s) (s - p) / s^2, lower = p, upper = xbar)$value
  atom_part <- (xbar - p) / xbar
  return(integral_part + atom_part)
}

cs_vals <- sapply(prices, compute_cs)
profit_vals <- rep(1, length(prices))

df_ph <- data.frame(price = prices, CS = cs_vals, profit = profit_vals)

ggplot(df_ph, aes(x = price, y = CS)) +
  geom_line(linewidth = 1, color = "darkgreen") +
  geom_point(data = data.frame(price = c(1, xbar), CS = c(0.6, 0)), 
             size = 3, color = "darkgreen") +
  scale_x_continuous("Price p", breaks = c(1, 1.4, xbar), 
                     labels = c("1", "1.4", "x̄")) +
  scale_y_continuous("Consumer Surplus", breaks = seq(0, 0.6, 0.1)) +
  theme_minimal() +
  ggtitle("Consumer Surplus vs Price (Profit = 1 throughout)")
```

## Part (i): All feasible (CS, profit) pairs

The feasible set is determined by the range of information structures for the buyer. By Theorem 2 from Lecture 6 (Roesler and Szentes, 2017), the feasible set is the *convex hull* of three points:

- $(0, p^*)$ where $p^* = 1$ is the minimum achievable profit
- $(\mu - p^*, p^*) = (0.6, 1)$ is the max-CS point
- $(0, \mu) = (0, 1.6)$ is the no-information point

*Key vertices:*

| Point | CS | Profit | Description |
|-------|-----|--------|-------------|
| A | 0 | 1.6 | No buyer information (seller extracts $E[v]$) |
| B | 0 | 1 | Minimum profit (truncated Pareto with $p = 1$) |
| C | 0.6 | 1 | Max-CS information structure (with $p = 1$) |

Note: Perfect buyer information gives $(0, 1.2)$, which lies *on the edge A–B*, not at a vertex.

*Boundary analysis:*

1. *Edge A–B (left edge):* CS $= 0$, profit $\in [1, 1.6]$. For any $\pi \in [1, 1.6]$, there exists a truncated Pareto $G_\pi^B$ with $\Pi(G_\pi^B) = \pi$. The seller can always set price $p = B$ (the upper bound of the support) to extract all surplus.

2. *Edge A–C (top edge):* Total surplus $= \mu = 1.6$ is achieved (no deadweight loss). So CS $+$ profit $= 1.6$, giving a line with slope $-1$.

3. *Edge B–C (bottom edge):* Profit $= p^* = 1$. With the truncated Pareto $G_1^B$ where $B = \bar{x} = e^{0.6}$, the seller is indifferent over prices $p \in [1, \bar{x}]$. Depending on which price is selected, CS ranges from $0$ (at $p = \bar{x}$) to $0.6$ (at $p = 1$).

```{r}
#| label: part-i
#| fig-cap: "All achievable (CS, Profit) pairs from buyer information structures"
#| warning: false
#| message: false

# Triangle vertices (corrected per Lecture 6)
dtriangle <- data.frame(
  cs = c(0, 0, 0.6, 0),
  profit = c(1.6, 1, 1, 1.6)
)

# Labelled points (vertices)
dpoints_i <- data.frame(
  cs = c(0, 0, 0.6),
  profit = c(1.6, 1, 1),
  label = c(
    "A: No buyer info\n(0, 1.6)",
    "B: Min profit\n(0, 1)",
    "C: Max CS\n(0.6, 1)"
  )
)

# Perfect information point (on edge, not vertex)
dpoint_perf <- data.frame(cs = 0, profit = 1.2)

ggplot() +
  geom_polygon(
    data = dtriangle, aes(x = cs, y = profit),
    fill = "steelblue", alpha = 0.25
  ) +
  geom_path(
    data = dtriangle, aes(x = cs, y = profit),
    linewidth = 0.8
  ) +
  geom_point(data = dpoints_i, aes(x = cs, y = profit), size = 3) +
  geom_point(data = dpoint_perf, aes(x = cs, y = profit), 
             size = 3, shape = 17, color = "darkred") +
  geom_text(
    data = dpoints_i, aes(x = cs, y = profit, label = label),
    vjust = c(-0.5, 1.5, -0.5), hjust = c(-0.1, -0.1, 0.5),
    size = 3.5
  ) +
  annotate("text", x = 0.08, y = 1.2, 
           label = "Perfect info\n(0, 1.2)", size = 3, hjust = 0) +
  annotate("text", x = 0.15, y = 1.35,
    label = "Feasible set", fontface = "italic", size = 4) +
  scale_x_continuous(
    "Consumer Surplus",
    limits = c(-0.05, 0.75),
    breaks = seq(0, 0.7, 0.1)
  ) +
  scale_y_continuous(
    "Profit",
    limits = c(0.6, 1.8),
    breaks = seq(0.6, 1.8, 0.2)
  ) +
  theme_minimal() +
  ggtitle("All (CS, Profit) pairs from buyer information structures")  
```

*Interpretation:*

The shaded triangle represents all achievable (consumer surplus, profit) pairs.

- *Less buyer information benefits the seller:* Moving up the left edge from B to A, the seller's profit increases from $1$ to $1.6$ as the buyer becomes less informed. Perfect buyer information ($\pi = 1.2$) lies strictly between these extremes.

- *The minimum profit is $p^* = 1$:* This is achieved by the truncated Pareto distribution from part (g). At this profit level, consumer surplus can range from $0$ to $0.6$ depending on the seller's price choice.

- *Trade-off structure:* The edge A–C represents efficient outcomes (no deadweight loss, total surplus $= 1.6$). Points strictly inside the triangle or on edge B–C involve some efficiency loss.

This result matches Theorem 2 from Lecture 6: the feasible set is the convex hull of $(0, p^*)$, $(\mu - p^*, p^*)$, and $(0, \mu)$.
