---
title: "Problem Set 7 - Solutions"
format: pdf
---

# Problem 1

Consider the following preferences:

- Alfred prefers Anna to Beth to Caroline to Deborah
- Ben prefers Caroline to Deborah to Anna to Beth
- Chris prefers Caroline to Anna to Deborah to Beth
- Dan prefers Deborah to Caroline to Beth to Anna
- Anna prefers Dan to Chris to Ben to Alfred
- Beth prefers Ben to Chris to Alfred to Dan
- Caroline prefers Alfred to Dan to Ben to Chris
- Deborah prefers Chris to Dan to Alfred to Ben

Check if the following matchings are stable:

a. Chris with Anna, Alfred with Beth, Dan with Caroline, Ben with Deborah
b. Chris with Anna, Alfred with Beth, Ben with Caroline, Dan with Deborah
c. Dan with Anna, Ben with Beth, Alfred with Caroline, Chris with Deborah

# Problem 2

For each of the following preferences, find both the woman-optimal and the man-optimal stable matchings.

(a)

|      | 1st  | 2nd  | 3rd  | 4th  | 
|------|------|------|------|------|
| Pat  | Lily | Hana | Mira | Rose |
| Jack | Hana | Mira | Rose | Lily |
| Karl | Mira | Rose | Lily | Hana |
| Les  | Hana | Lily | Mira | Rose |

|      | 1st  | 2nd  | 3rd  | 4th |
|------|------|------|------|-----|
| Hana | Karl | Pat  | Jack | Les |
| Lily | Jack | Karl | Pat  | Les |
| Mira | Pat  | Jack | Karl | Les |
| Rose | Pat  | Jack | Karl | Les |

(b)

|       | 1st  | 2nd         | 3rd         |
|-------|------|-------------|-------------|
| Alan  | Sara | Rose        | $\emptyset$ |
| Basil | Rose | $\emptyset$ | Sara        |
| Colin | Rose | Sara        | $\emptyset$ |

|      | 1st   | 2nd   | 3rd         | 4th   |
|------|-------|-------|-------------|-------|
| Rose | Alan  | Basil | $\emptyset$ | Colin |
| Sara | Basil | Colin | $\emptyset$ | Alan  |

# Solution to Problem 1

A matching is *stable* if:

1. It is *individually rational*: no one prefers being unmatched to their current partner
2. There is *no blocking pair*: no $(m, w)$ such that $w \succ_m \mu(m)$ and $m \succ_w \nu(w)$

```{r}
#| label: setup
#| warning: false
#| message: false

library(ggplot2)
library(dplyr)
library(tidyr)
```

Let's encode the preferences for reference:

```{r}
#| label: preferences
#| warning: false
#| message: false

# Men's preferences (1 = most preferred)
men_prefs <- data.frame(
  man = c("Alfred", "Ben", "Chris", "Dan"),
  rank1 = c("Anna", "Caroline", "Caroline", "Deborah"),
  rank2 = c("Beth", "Deborah", "Anna", "Caroline"),
  rank3 = c("Caroline", "Anna", "Deborah", "Beth"),
  rank4 = c("Deborah", "Beth", "Beth", "Anna")
)

# Women's preferences (1 = most preferred)
women_prefs <- data.frame(
  woman = c("Anna", "Beth", "Caroline", "Deborah"),
  rank1 = c("Dan", "Ben", "Alfred", "Chris"),
  rank2 = c("Chris", "Chris", "Dan", "Dan"),
  rank3 = c("Ben", "Alfred", "Ben", "Alfred"),
  rank4 = c("Alfred", "Dan", "Chris", "Ben")
)
```

## Part (a): Chris-Anna, Alfred-Beth, Dan-Caroline, Ben-Deborah

Let's check for blocking pairs systematically:

| Pair | Does she prefer him? | Does he prefer her? | Blocking? |
|------|---------------------|---------------------|-----------|
| (Alfred, Anna) | Alfred is 4th, current (Chris) is 2nd. No | - | No |
| (Alfred, Caroline) | Alfred is 1st, current (Dan) is 2nd. *Yes* | Caroline is 3rd, current (Beth) is 2nd. No | No |
| (Alfred, Deborah) | Alfred is 3rd, current (Ben) is 4th. *Yes* | Deborah is 4th, current (Beth) is 2nd. No | No |
| (Ben, Anna) | Ben is 3rd, current (Chris) is 2nd. No | - | No |
| (Ben, Beth) | Ben is 1st, current (Alfred) is 3rd. *Yes* | Ben: Beth is 4th, current (Deborah) is 2nd. No | No |
| (Ben, Caroline) | Ben is 3rd, current (Dan) is 2nd. No | - | No |
| (Chris, Beth) | Chris is 2nd, current (Alfred) is 3rd. *Yes* | Beth is 4th, current (Anna) is 2nd. No | No |
| (Chris, Caroline) | Chris is 4th, current (Dan) is 2nd. No | - | No |
| (Chris, Deborah) | Chris is 1st, current (Ben) is 4th. *Yes* | Deborah is 3rd, current (Anna) is 2nd. No | No |
| (Dan, Anna) | Dan is 1st, current (Chris) is 2nd. *Yes* | Anna is 4th, current (Caroline) is 2nd. No | No |
| (Dan, Beth) | Dan is 4th, current (Alfred) is 3rd. No | - | No |
| (Dan, Deborah) | Dan is 2nd, current (Ben) is 4th. *Yes* | Deborah is 1st, current (Caroline) is 2nd. *Yes* | *YES* |

$$\text{Matching (a) is NOT stable: (Dan, Deborah) is a blocking pair}$$

Dan prefers Deborah (his 1st choice) to Caroline (his current match, 2nd choice).
Deborah prefers Dan (her 2nd choice) to Ben (her current match, 4th choice).

## Part (b): Chris-Anna, Alfred-Beth, Ben-Caroline, Dan-Deborah

Let's check for blocking pairs:

| Pair | Does she prefer him? | Does he prefer her? | Blocking? |
|------|---------------------|---------------------|-----------|
| (Alfred, Anna) | Alfred is 4th, Chris is 2nd. No | - | No |
| (Alfred, Caroline) | Alfred is 1st, Ben is 3rd. *Yes* | Caroline is 3rd, Beth is 2nd. No | No |
| (Alfred, Deborah) | Alfred is 3rd, Dan is 2nd. No | - | No |
| (Ben, Anna) | Ben is 3rd, Chris is 2nd. No | - | No |
| (Ben, Beth) | Ben is 1st, Alfred is 3rd. *Yes* | Ben: Beth is 4th, Caroline is 1st. No | No |
| (Ben, Deborah) | Ben is 4th, Dan is 2nd. No | - | No |
| (Chris, Beth) | Chris is 2nd, Alfred is 3rd. *Yes* | Beth is 4th, Anna is 2nd. No | No |
| (Chris, Caroline) | Chris is 4th, Ben is 3rd. No | - | No |
| (Chris, Deborah) | Chris is 1st, Dan is 2nd. *Yes* | Deborah is 3rd, Anna is 2nd. No | No |
| (Dan, Anna) | Dan is 1st, Chris is 2nd. *Yes* | Anna is 4th, Deborah is 1st. No | No |
| (Dan, Beth) | Dan is 4th, Alfred is 3rd. No | - | No |
| (Dan, Caroline) | Dan is 2nd, Ben is 3rd. *Yes* | Caroline is 2nd, Deborah is 1st. No | No |

$$\text{No blocking pairs found} \implies \text{Matching (b) is stable}$$

## Part (c): Dan-Anna, Ben-Beth, Alfred-Caroline, Chris-Deborah

Let's check for blocking pairs:

| Pair | Does she prefer him? | Does he prefer her? | Blocking? |
|------|---------------------|---------------------|-----------|
| (Alfred, Anna) | Alfred is 4th, Dan is 1st. No | - | No |
| (Alfred, Beth) | Alfred is 3rd, Ben is 1st. No | - | No |
| (Alfred, Deborah) | Alfred is 3rd, Chris is 1st. No | - | No |
| (Ben, Anna) | Ben is 3rd, Dan is 1st. No | - | No |
| (Ben, Caroline) | Ben is 3rd, Alfred is 1st. No | - | No |
| (Ben, Deborah) | Ben is 4th, Chris is 1st. No | - | No |
| (Chris, Anna) | Chris is 2nd, Dan is 1st. No | - | No |
| (Chris, Beth) | Chris is 2nd, Ben is 1st. No | - | No |
| (Chris, Caroline) | Chris is 4th, Alfred is 1st. No | - | No |
| (Dan, Beth) | Dan is 4th, Ben is 1st. No | - | No |
| (Dan, Caroline) | Dan is 2nd, Alfred is 1st. No | - | No |
| (Dan, Deborah) | Dan is 2nd, Chris is 1st. No | - | No |

$$\text{No blocking pairs found} \implies \text{Matching (c) is stable}$$

```{r}
#| label: matching-comparison
#| fig-cap: "Summary of stable matchings for Problem 1"
#| warning: false
#| message: false

# Visualize the matchings
matchings <- data.frame(
  matching = rep(c("(a)", "(b)", "(c)"), each = 4),
  man = rep(c("Alfred", "Ben", "Chris", "Dan"), 3),
  woman = c(
    "Beth", "Deborah", "Anna", "Caroline",  # (a)
    "Beth", "Caroline", "Anna", "Deborah",  # (b)
    "Caroline", "Beth", "Deborah", "Anna"   # (c)
  ),
  stable = rep(c("Not Stable", "Stable", "Stable"), each = 4)
)

ggplot(matchings, aes(x = man, y = woman, fill = stable)) +
  geom_tile(color = "white", linewidth = 1) +
  facet_wrap(~matching) +
  scale_fill_manual(values = c("Stable" = "steelblue", "Not Stable" = "tomato")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Men", y = "Women", fill = "Status") +
  ggtitle("Matchings from Problem 1")
```

# Solution to Problem 2

## Part (a): 4x4 Marriage Market

*Preferences:*

- Pat: Lily $\succ$ Hana $\succ$ Mira $\succ$ Rose
- Jack: Hana $\succ$ Mira $\succ$ Rose $\succ$ Lily  
- Karl: Mira $\succ$ Rose $\succ$ Lily $\succ$ Hana
- Les: Hana $\succ$ Lily $\succ$ Mira $\succ$ Rose

- Hana: Karl $\succ$ Pat $\succ$ Jack $\succ$ Les
- Lily: Jack $\succ$ Karl $\succ$ Pat $\succ$ Les
- Mira: Pat $\succ$ Jack $\succ$ Karl $\succ$ Les
- Rose: Pat $\succ$ Jack $\succ$ Karl $\succ$ Les

### Man-Optimal Stable Matching (Men Propose)

*Round 1:*

- Pat proposes to Lily (his 1st choice)
- Jack proposes to Hana (his 1st choice)
- Karl proposes to Mira (his 1st choice)
- Les proposes to Hana (his 1st choice)

Hana receives proposals from Jack and Les. She prefers Jack (3rd) to Les (4th).
*Les is rejected.*

*Tentative matching:* Pat-Lily, Jack-Hana, Karl-Mira

*Round 2:*

- Les proposes to Lily (his 2nd choice)

Lily receives proposals from Pat and Les. She prefers Pat (3rd) to Les (4th).
*Les is rejected.*

*Round 3:*

- Les proposes to Mira (his 3rd choice)

Mira receives proposals from Karl and Les. She prefers Karl (3rd) to Les (4th).
*Les is rejected.*

*Round 4:*

- Les proposes to Rose (his 4th choice)

Rose accepts Les.

*Man-optimal stable matching:*

$$\text{Pat-Lily, Jack-Hana, Karl-Mira, Les-Rose}$$

### Woman-Optimal Stable Matching (Women Propose)

*Round 1:*

- Hana proposes to Karl (her 1st choice)
- Lily proposes to Jack (her 1st choice)
- Mira proposes to Pat (her 1st choice)
- Rose proposes to Pat (her 1st choice)

Pat receives proposals from Mira and Rose. He prefers Mira (3rd) to Rose (4th).
*Rose is rejected.*

*Tentative matching:* Hana-Karl, Lily-Jack, Mira-Pat

*Round 2:*

- Rose proposes to Jack (her 2nd choice)

Jack receives proposals from Lily and Rose. He prefers Rose (3rd) to Lily (4th).
*Lily is rejected.*

*Tentative matching:* Hana-Karl, Rose-Jack, Mira-Pat

*Round 3:*

- Lily proposes to Karl (her 2nd choice)

Karl receives proposals from Hana and Lily. He prefers Lily (3rd) to Hana (4th).
*Hana is rejected.*

*Tentative matching:* Lily-Karl, Rose-Jack, Mira-Pat

*Round 4:*

- Hana proposes to Pat (her 2nd choice)

Pat receives proposals from Mira and Hana. He prefers Hana (2nd) to Mira (3rd).
*Mira is rejected.*

*Tentative matching:* Lily-Karl, Rose-Jack, Hana-Pat

*Round 5:*

- Mira proposes to Jack (her 2nd choice)

Jack receives proposals from Rose and Mira. He prefers Mira (2nd) to Rose (3rd).
*Rose is rejected.*

*Tentative matching:* Lily-Karl, Mira-Jack, Hana-Pat

*Round 6:*

- Rose proposes to Karl (her 3rd choice)

Karl receives proposals from Lily and Rose. He prefers Rose (2nd) to Lily (3rd).
*Lily is rejected.*

*Tentative matching:* Rose-Karl, Mira-Jack, Hana-Pat

*Round 7:*

- Lily proposes to Pat (her 3rd choice)

Pat receives proposals from Hana and Lily. He prefers Lily (1st) to Hana (2nd).
*Hana is rejected.*

*Tentative matching:* Rose-Karl, Mira-Jack, Lily-Pat

*Round 8:*

- Hana proposes to Jack (her 3rd choice)

Jack receives proposals from Mira and Hana. He prefers Hana (1st) to Mira (2nd).
*Mira is rejected.*

*Tentative matching:* Rose-Karl, Hana-Jack, Lily-Pat

*Round 9:*

- Mira proposes to Karl (her 3rd choice)

Karl receives proposals from Rose and Mira. He prefers Mira (1st) to Rose (2nd).
*Rose is rejected.*

*Tentative matching:* Mira-Karl, Hana-Jack, Lily-Pat

*Round 10:*

- Rose proposes to Les (her 4th choice)

Les accepts Rose (she's his 4th choice, but better than being unmatched).

*Woman-optimal stable matching:*

$$\text{Pat-Lily, Jack-Hana, Karl-Mira, Les-Rose}$$

The plot's numbers show partner's rank (M = man's rank for her, W = woman's rank for him)

In this example, the man-optimal and woman-optimal stable matchings are identical. This means the stable matching is unique.

```{r}
#| label: part-2a
#| fig-cap: "DAA for Problem 2(a) - Unique stable matching"
#| warning: false
#| message: false

# The unique stable matching
matching_2a <- data.frame(
  man = c("Pat", "Jack", "Karl", "Les"),
  woman = c("Lily", "Hana", "Mira", "Rose"),
  man_rank = c(1, 1, 1, 4),  # Their partner's rank in their preference
  woman_rank = c(3, 3, 3, 4)  # Their partner's rank in their preference
)

ggplot(matching_2a, aes(x = man, y = woman)) +
  geom_tile(fill = "steelblue", alpha = 0.3, color = "steelblue", linewidth = 1) +
  geom_text(aes(label = paste0("M:", man_rank, " W:", woman_rank)), size = 4) +
  theme_minimal() +
  labs(x = "Men", y = "Women") +
  ggtitle("Problem 2(a): Unique Stable Matching")
```

## Part (b): 3x3 Market with Unacceptable Partners

*Preferences (with $\emptyset$ indicating unacceptable):*

- Alan: Sara $\succ$ Rose $\succ$ $\emptyset$
- Basil: Rose $\succ$ $\emptyset$ $\succ$ Sara
- Colin: Rose $\succ$ Sara $\succ$ $\emptyset$

- Rose: Alan $\succ$ Basil $\succ$ $\emptyset$ $\succ$ Colin
- Sara: Basil $\succ$ Colin $\succ$ $\emptyset$ $\succ$ Alan

Note: Colin is unacceptable to Rose, and Alan is unacceptable to Sara.

### Man-Optimal Stable Matching (Men Propose)

*Round 1:*

- Alan proposes to Sara (his 1st choice)
- Basil proposes to Rose (his 1st choice)
- Colin proposes to Rose (his 1st choice)

Sara rejects Alan (he's unacceptable to her).
Rose receives proposals from Basil and Colin. Colin is unacceptable to her, so she considers only Basil. *Rose tentatively accepts Basil, Colin is rejected.*

*Tentative matching:* Basil-Rose

*Round 2:*

- Alan proposes to Rose (his 2nd choice)
- Colin proposes to Sara (his 2nd choice)

Rose receives proposals from Basil (current) and Alan. She prefers Alan (1st) to Basil (2nd).
*Basil is rejected.*

Sara accepts Colin (he's her 2nd choice, acceptable).

*Tentative matching:* Alan-Rose, Colin-Sara

*Round 3:*

- Basil proposes to $\emptyset$ (his 2nd choice) — remains unmatched

Sara is unacceptable to Basil, so he prefers staying single.

*Man-optimal stable matching:*

$$\text{Alan-Rose, Colin-Sara, Basil unmatched}$$

### Woman-Optimal Stable Matching (Women Propose)

*Round 1:*

- Rose proposes to Alan (her 1st choice)
- Sara proposes to Basil (her 1st choice)

Alan accepts Rose (his 2nd choice).
Basil rejects Sara (she's unacceptable to him).

*Tentative matching:* Rose-Alan

*Round 2:*

- Sara proposes to Colin (her 2nd choice)

Colin accepts Sara (his 2nd choice).

*Tentative matching:* Rose-Alan, Sara-Colin

*Woman-optimal stable matching:*

$$\text{Alan-Rose, Colin-Sara, Basil unmatched}$$

Again, the stable matching is unique.

```{r}
#| label: part-2b
#| fig-cap: "Problem 2(b) - Market with unacceptable partners"
#| warning: false
#| message: false

# The stable matching (with Basil unmatched)
matching_2b <- data.frame(
  man = c("Alan", "Basil", "Colin"),
  woman = c("Rose", NA, "Sara"),
  status = c("Matched", "Unmatched", "Matched")
)

ggplot(matching_2b, aes(x = man, y = woman, fill = status)) +
  geom_tile(color = "white", linewidth = 1) +
  geom_text(aes(label = ifelse(status == "Matched", "M", "U")), size = 8) +
  scale_fill_manual(values = c("Matched" = "steelblue", "Unmatched" = "gray80")) +
  theme_minimal() +
  labs(x = "Men", y = "Women/Status", fill = "") +
  ggtitle("Problem 2(b): Unique Stable Matching")
```

## Summary

```{r}
#| label: summary-table
#| warning: false
#| message: false

summary_df <- data.frame(
  Problem = c("2(a)", "2(a)", "2(b)", "2(b)"),
  Algorithm = c("Man-optimal (men propose)", "Woman-optimal (women propose)",
                "Man-optimal (men propose)", "Woman-optimal (women propose)"),
  Matching = c(
    "Pat-Lily, Jack-Hana, Karl-Mira, Les-Rose",
    "Pat-Lily, Jack-Hana, Karl-Mira, Les-Rose",
    "Alan-Rose, Colin-Sara, Basil-NA",
    "Alan-Rose, Colin-Sara, Basil-NA"
  )
)

knitr::kable(summary_df, caption = "Summary of stable matchings")
```

*Key insights from the lectures:*

1. *Theorem 1 (Existence):* The Deferred Acceptance Algorithm always produces a stable matching.

2. *Theorem 2 (Side-Optimality):* The DAA where one side proposes produces the optimal stable matching for that side. Every proposer gets their best stable partner, while every receiver gets their worst stable partner.

3. *Lattice Structure:* The set of stable matchings forms a complete lattice. When the man-optimal and woman-optimal matchings coincide (as in both problems above), the stable matching is unique.
